The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-16 paper/ms.tex
Fix travis
Converted this text to an issue
Add citation to new GRAVITY measurement
Update citation to Trick et al paper
Add section of analytic action errors in epic. approx.

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{bm}

@@

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center is taken to be $\sim 8.1\text{--}8.3\,\kpc$
\citep{2009ApJ...692.1075G,2018AA...615L..15G}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

@@

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of {\em Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B,2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2018arXiv180503653T}.

@@

@@

Other methods for computing actions are used in the literature. For example,
the St\"ackel Fudge method \citep{2016MNRAS.457.2107S}, which uses a single
St\"ackel potential (with analytic actions) to approximate the Galactic
potential \citep{1985MNRAS.216..273D,2012MNRAS.426.1324B}, was used in many
recent works exploring actions in the Galactic disk
\citep[e.g.][]{2018arXiv180503653T,2018MNRAS.481.4093S,2018arXiv180803278T}.
For disk-like orbits, existing implementations of the St\"ackel Fudge method
are of acceptable accuracy, but since we also consider halo-like orbits in
this work (where the St\"ackel Fudge method is inaccurate) we choose to use
orbit integration and torus mapping throughout \citep{2016MNRAS.457.2107S}.

@@

%%%move this later - decide where, probably conclusions While we focus our
%discussion now solely on the midplane, there is an additional complication in
%the measurement of the distance from the Sun to the Galactic center that may
%also be important. When modelling stars at Galactic radii close to the Sun's
%radius, it is in principle possible to determine what point they are orbiting
%about: the ``dynamical Galactic center.'' It is not guaranteed that this
%dynamical center will coincide with the distance to Sgr~A\textsuperscript{*}
%\citep{2018AA...615L..15G}. In fact, \citet{2015ApJ...803...80K} were able to
%measure the distance of the Sun to the Galactic center using the dynamical
%information of Palomar 5. Other dynamical measurements of the Solar distance
%have been made, though none with a precision capable of competing with the
%distance to Sgr~A\textsuperscript{*}
%\citep{1981gask.book.....M,2011PASJ...63..867S,2012MNRAS.427..274S,2013AstL...39...95B,2013IAUS..289..444Z}.

>>>>

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{bm}
\usepackage{mathtools}

@@

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center is taken to be $\sim 8.1\text{--}8.3\,\kpc$
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent $\sim0.3\%$ measurement from the GRAVITY collaboration \citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

@@

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of {\em Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B,2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2019MNRAS.484.3291T}.

@@

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the
induced action error from either a midplane offset or a Galactocentric radius
offset. We make the epicyclic approximation, which assumes that the motion in
the $z$ and $R$ components of the orbit are decoupled and follow a simple
harmonic oscillator. This approximation should be an excellent description of
thin disk-like orbits and a good description of thick disk-like orbits. We
also make the assumption of a perfectly flat circular velocity curve with
velocity $v_c$.

Under this approximation, we can write down the vertical and radial components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
z(t) &= A_z \sin{(\kappa t + \alpha)}\\
R(t) &= A_R \sin{(\nu t + \beta)}\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. We can write down the actions in this approximation as
\citep[Section~3.5.3b][]{2008gady.book.....B},
\beq\label{eq:actions_epi}
\begin{split}
L_z &= R_g v_c \\
J_r &= \kappa A_R^2 \\
J_z &= \nu A_z^2\text{,}
\end{split}
\eeq
where $R_g$ is the guiding radius of the orbit.

An induced error in $z$ of $\Delta z$ will cause $\Delta A_z \sim \Delta z$.
Similarly, an induced error in $R$ will cause $\Delta A_R \sim \Delta R$. If
we assume that $\kappa$ and $\nu$ are constant (\Gus{justify?}), then we have
that
\beq\label{eq:actions_epi_error}
\begin{split}
\Delta L_z = \Delta R v_c &\implies \frac{\Delta L_z}{L_z} = \frac{\Delta R}{R} \\
\Delta J_r = 2 \kappa \Delta A_R &\implies \frac{\Delta J_r}{J_r} = \frac{2\Delta R}{A_R} \\
\Delta J_z = 2 \nu \Delta A_z &\implies \frac{\Delta J_z}{J_z} = \frac{2\Delta z}{A_z}\text{.}
\end{split}
\eeq
Therefore, errors in each action are linearly proportional to the errors in
either the midplane or cylindrical radius. We will check these formulae
against numerical tests next.

@@

Other methods for computing actions are used in the literature. For example,
the St\"ackel Fudge method \citep{2016MNRAS.457.2107S}, which uses a single
St\"ackel potential (with analytic actions) to approximate the Galactic
potential \citep{1985MNRAS.216..273D,2012MNRAS.426.1324B}, was used in many
recent works exploring actions in the Galactic disk
\citep[e.g.][]{2019MNRAS.484.3291T,2018MNRAS.481.4093S,2018arXiv180803278T}.
For disk-like orbits, existing implementations of the St\"ackel Fudge method
are of acceptable accuracy, but since we also consider halo-like orbits in
this work (where the St\"ackel Fudge method is inaccurate) we choose to use
orbit integration and torus mapping throughout \citep{2016MNRAS.457.2107S}.

@@
